---
title: "Homework 5"
author: "Kristi Gdanetz MacCready"
date: "02/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMBio201/Lab5/")
```

# Load packages
```{r Load packages, message=FALSE, warning=FALSE, include=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
set.seed(7)
```

Save any data frames generated to curated_data directory, save all plots to figures directory. Include any statistics (p- or R-values) as comments after the tests. Based on assumptions, clearly state which statistical test will be used and why. 

# Question 1

Use the SCFA wide dataframe from Lab 5. Check assumptions, conduct statistical tests, make a plot, to answer the question, "Which supplement causes the greatest increase in acetate concentrations?"
```{r}
# import and filter data set (q1_df)
```

```{r}
# check assumptions
```

```{r}
# conduct statistical test
```

```{r}
# make and save plot (q1_plot)
```

What are your conclusions regarding the reseach question? 

<type answer here>


# Question 2

Use the SCFA wide dataframe from Lab 5. Check assumptions, conduct statistical tests, make a plot, to answer the question, "Which supplement causes the greatest increase in propionate concentrations?"
```{r}
# import and filter data set (q2_df)
```

```{r}
# check assumptions
```

```{r}
# conduct statistical test
```

```{r}
# make and save plot (q2_plot)
```

What are your conclusions regarding the reseach question?

<type answer here>


# Question 3

Use the SCFA wide dataframe from Lab 5. Subset and filter the data frame, check assumptions, and choose the appropriate statistical tests, and create a plot to answer the following research question: "Does the proportion of acetate reponses differ between individuals who consumed BRMPS potato (BRMPS) or LOODAT potato supplements?"

```{r}
# import and filter data set (q3_df)
```

```{r}
# check assumptions
```

```{r}
# conduct statistical test
```

```{r}
# make and save plot (q3_plot)
```

What are your conclusions regarding the reseach question?

<type answer here>


# Question 4

Use the SCFA wide dataframe from Lab 5. Subset and filter the data frame, check assumptions, and choose the appropriate statistical tests, and create a plot to answer the following research question: "Does the proportion of propionate reponses differ between individuals who consumed BRMPS potato (BRMPS) or LOODAT potato supplements?"

```{r}
# import and filter data set (q4_df)
```

```{r}
# check assumptions
```

```{r}
# conduct statistical test
```

```{r}
# make and save plot (q4_plot)
```

What are your conclusions regarding the reseach question?

<type answer here>


-----
end